Determination of Center of Rotation \\ in Computed Tomography

### 2017-07-26 seminar_cor/seminar_cor_haegele.tex
added presentation

% The paper headers
\markboth{Seminar on Image Processing - Algorithms and Architectures - Univerity of Stuttgart - IPVS/PAS - SS17}%
{H{\"a}gele: Determination of Center of Rotation in Computed Tomography}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
\begin{abstract}
This paper will discuss the aspect of the determination of center of rotation in computed tomography.
An introduction to computed tomography imaging will be given in the beginning as a prerequisite to the actual topic. 
This will cover different geometries (parallel beam, fan beam and cone beam) which are present in computed tomography and used to model the projection space of a CT-scan as well as basic knowledge about tomographic reconstruction, including the Fourier slice theorem and filtered back projection.
The problem of wrong center of rotation assumption in tomographic reconstruction will be explained which has the effect of image degradation.
Different solutions to the determination of the correct location of the center of rotation found in the literature will be presented.
It is found that the approaches can be split into sinogram and reconstruction based methods, where sinogram based methods analyze the sinogram that is the product of a CT-scan, while reconstruction based methods analyze reconstructed images of different centers of rotation to find the correct one.
Finally a conclusion summarizing the paper will be drawn.  
\end{abstract}

@@

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{I}{n} computed tomography (CT), a set of set of projections is acquired by meassuring the attenuation of \xrays through the scanned object at different rotation angles.
The \xrays are emmited by a source on one side of the scanned object, intersect with the object, and hit a detector on the other side.
The resulting meassurement of the detector is a projection of the object's internals.
Rotating the object and repeating the process results in a set of projections, which can be used to reconstruct an image of the internal structure of the scanned object. 
For this reconstruction it is mandatory to know the center of rotation (COR), or rather the location of its projection, as slight deviations from the actual COR produce severe artifacts in the reconstructed image.
In general, the COR is assumed to be projected onto the center of the detector, but this assumption can be violated on badly calibrated CT systems. 
A correct calibration can also become almost impossible when the \xray source is very close to the scanned object, as is the case for industrial scanners with resolutions on the micrometer scale \cite{azevedo90}.
In this paper, different techniques and algorithms for the determination of the center of rotation are explained, which can be found in the literature.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an

@@

\section{Prerequisites}
In order to explain the problem of inaccurate COR assumption in tomographic reconstruction, some prerquisite knowledge about CT will be briefly explained in this section.

\subsection{Image Acquisition in CT}\label{ssec:imageacquisition}
This subsection will briefly explain how images are typically aquired in CT.

@@

\subsubsection{Parallel Beam}\label{sssec:parallelbeam}
The simplest geometry in CT (from a mathematical point of view) is the parallel beam geometry.
Here all rays are parallel that are emitted from the same side of the object.
\Cref{subfig:parallelbeam} shows a schematic of this geometry with the resulting signal.
%
\begin{figure}[!h]
\centering
\subfloat[Parallel Beam]{\includegraphics[width=0.58\linewidth]{img/parallel_beam_geometry}%
\label{subfig:parallelbeam}}
\hfil
\subfloat[Fan Beam]{\includegraphics[width=0.4\linewidth, trim=0 -6cm 0 0]{img/fanbeamgeometry}%
\label{subfig:fanbeam}}
\caption{Parallel beam and fan beam geometry. 
In parallel beam the source moving perpendicular to the ray direction for sampling a 1-dimensional image (graph). 
In fan beam, rays are not perpendicular but instead emmited in fan blade form. 
$\gamma$ is the rotation angle at which the signal is obtained. 
Figures from~\cite{Buzug2008_chap3,Buzug2008_chap4}}
\label{fig:parallalandfanbeam}
\end{figure}

@@

Unfortuantely, parallel beam geometry is impractical as a pencil beam \xray source (source emmiting only one ray in a specific direction) would have to be moved parallel to the detector to collect all samples for a single rotation angle, which is very time consuming \cite{Buzug2008_chap7}.
However, parallel beam geometry occurs at synchrotron facilities \cite{something}.

\subsubsection{Fan Beam}\label{sssec:fanbeam}
A more practical geometry is the fan beam, in which a source emits a flat beam that extends to the sides (like a fan blade). 
\Cref{subfig:fanbeam} shows a schematic of this geometry.
Here the projection for a single rotation angle can be obtained at once. 
The geometry can be defined by the COR it rotates around, the offset of the source to the COR and its aperture angle.
In clinical practise curved detector rays are used, as shown in \cref{subfig:fanbeam}, to get constant aperture angle steps for each detector cell.
%
\begin{figure}[!h]
\centering
\subfloat[]{\includegraphics[width=0.43\linewidth]{eps/rebinning.eps}%
\label{subfig:rebinning1}}
\hfil
\subfloat[]{\includegraphics[width=0.51\linewidth]{eps/rebinning_geo.eps}%
\label{subfig:rebinning2}}
\caption{
On the left a schematic is shown with two fan beams with aperture angle $\varphi$ at different rotation angles.
It can be sen that the outer rays of the two are parallel.
The maximum distance of two parallel beams is limited by $\varphi$, and the maximum rotation angle $\theta$ at which two
parallel rays can be obtained is given by $\theta = \psi_{min}+\psi_{max}$ which are the angles from central to outer rays.
On the right a schematic is shown that points out the correspondence between fan and parallel beam geometry.
It can be seen that for the red ray which is the ray at rotation angle $\theta$ with deviation angle $\psi$ from the central ray, there exists an angle $\gamma$ in parallel beam geometry so that it is perpendicular to the $r$-axis (i.e. parallel beam detector).
Figures from~\cite{Buzug2008_chap7}
}
\label{fig:rebinning}
\end{figure}
%

@@

As th Radon transform is a set of projections the Fourier slice theorem can be used to build the Fourier tranform of the original signal and reconstruct it using the inverse Fourier transform.
\Cref{fig:inverseradon} illustrates this method.
%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{img/inverseradon.png}
\caption{
The inverse Radon transform can be calculated using the Fourier slice theorem.
First all projections are transformed to the Fourier space along the $\xi$-axis ($=r$-axis).
Then the Fourier transform is built by placing all transformed projections according to their angle into a polar coordinate system.
The coordinate system thus consists of slices of the Fourier transform of the original signal which can then
be obtained using an inverse Fourier transform.
Figure from \cite{Buzug2008_chap5}
}
\label{fig:inverseradon}
\end{figure}
%
The problem of applying this algorithm in practise is, that due to the limited angular resolution (i.e. discretization) of the Radon transform the frequency information density in the Fourier space decreases for higher frequencies \cite{Buzug2008_chap5}.
In the process of cartesian regridding of the polar coordinates a lot of data for the high frequencies would have to be made up in order to perfrom an inverse fast fourier transform.
Therefore a different algorithm, the filtered back projection, is used in practise.

@@

This can be imagined as smearing the projections over a canvas in their respective directions.
At first glance this may seem as the correct inversion of the projections but is not.
This becomes clear when thinking about the original image being a dirac delta (image with all entreis zero except for the central pixel).
The projections of this image will be 1D dirac deltas, but using the back projection algorithm will not
restore the 2D dirac but instead smear the 1D diracs also into areas which are supposed to be zero.
In fact the resulting image looks quite blurry which is due to the back projection equaling to the original image convolved with the point spread function of the imaging system (i.e. the Radon transform) $\frac{1}{|(x,y)|}$ \cite{Buzug2008_chap5}.
To counter this, the projections first need to be filtered with a high pass filter.
This introduces negative values in the projections which in the back projection process cancel out positive contributions of other projections.
Figure \cref{fig:backprojection} shows the results of unfiltered and filtered back projection.
As amplifying high frequencies also aplifies noise, there are different filters that can be used in practise \cite{farquhar1998}.
%
\begin{figure}[!h]
\centering
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_02}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_04}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_08}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_16}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_32}}
\\
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_02}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_04}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_08}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_16}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_32}}
\caption{
On the top row, the results of unfiltered back projection are shown. 
From left to right the number of projections used increases from 2 to 32.
The resulting image looks blurry.
On the bottom row the results of filtered back projection are shown.
The resulting image looks quite crips and edges are clearly visible.
The Shepp Logan phantom~\cite{shepplogan} was used as source image.  
}
\label{fig:backprojection}
\end{figure}
%

@@

\subsection{Sinogram Based Methods}
In this subsection the approaches that are based on sinogram analysis are introduced.
The approaches are presentred in chronical order.

@@

This approach could also be expanded to use all projections in the sinogram.
For each opposing pair of projections the COR could be calcualted and then the mean of all CORs will yield the final result.
This approach can be automated if there esists a distinct global maximum in the projections of the sinogram, otherwise a location of the correct feature is not guaranteed.

@@

Azevedo~et~al. showed that their algorithm is quite robust against Gaussian noise and isolated singularities~\cite{azevedo90}.
Unfortuanately their algorithm is not applicable to fan beam geometry in general the objects center of mass is no longer projected onto the projections center of mass as was the case for parallel beam.
However for an object function $f(x,y) = \delta(x-\overline{x}, y-\overline{y})$, that is a dirac delta function translated by $(\overline{x},\overline{y})$, the center of mass is projected onto the projections center of mass.
Therefore this method can be used for calibration (i.e. finding the COR beforehand) using an object that approximates a dirac delta like a thin wire or pin phantom.

\subsubsection{Cross-Correlation}\label{sssec:crosscorrelation}
Yang et al.~\cite{Yang2011} proposed another sinogram based method for COR determination using image cross-correllation.
Given a sinogram in a prallel beam geometry over $360^\circ$ (which can also be obtained by rebinning a fan-beam sinogram), the sinogram is divided in half so that both halfs cover a $180^\circ$ range.
As the information in both sub sinograms is redundant in that way that they are mirror images of each other,
flipping one of them will result in two sinograms over $180^\circ$ with identical information.
%
\begin{equation}
R(r,\gamma)^{[0,\pi]} = R(-r,\gamma)^{[\pi,2\pi]}
\end{equation}
%
A shift of COR, as explained in the beginning of \cref{sec:cordetermination}, results in a shift of the sinogram along the $r$-axis (see~\cref{eq:radonoffset}).
Thus, the two sub sinograms are identical under this shift.
%
\begin{equation}
R_\tau(r+\tau,\gamma)^{[0,\pi]} = R_\tau(-(r+\tau),\gamma)^{[\pi,2\pi]}
\end{equation}
%
Using the 1D cross correlation of the two sinograms along the $r$-axis, the shift can be determined as follows.
%
\begin{equation}
2\tau = \underset{r}{arg\,max}\quad R_\tau(r,\gamma)^{[0,\pi]} \star R_\tau(-r,\gamma)^{[\pi,2\pi]}
\end{equation}
%
The authors~\cite{Yang2011} provided evidence that their algorithm is robust against Gaussian noise even for large variance values.
They also stated that their algorithm can be used for 3D CT with cone beam geometry when using the central slice (see \cref{sssec:conebeam}).
As locating the central slice is not trivial and depends on calibration as well, it was shown that the algorithm still performs well for sinograms that slightly deviate from that of the central slice.

@@

Recalling the filtering step in FBP where negative projection values are introduced to cancel out positive contribution (see~\cref{sssec:fbp}), it makes sense to analyse the negativity of the reconstructed image.
In a perfect reconstruction the negativity would be zero as the original image had no negative values, so minimizing this metric yields the correct COR.

@@

\subsubsection{Histogram Entropy}\label{sssec:qh}
As pointed out by Donath et~al.~\cite{donath2006}, the metric $Q_{IN}$ can only be used for non-negative object functions.
Therefore they provided another metric based on the histogram entropy that is also applicable when negative attenuation coefficients are present (which can occur when samples are recorded in a surrounding medium e.g. water \cite{donath2006}).
They defined a metric $Q_H$ as follows.
%
\begin{align}
Q_H(\tilde{f}) &= \frac{H}{H_{max}} 
\\
H &= -\int_{-\infty}^{\infty}p(g)\,log_2(h\,p(g))\;dg
\end{align}
%
Here the metric is build by normalizing the entropy $H$ by the maximum entropy $H_{max}$.
$H$ is a continous version of the discrete entropy for gray-level images, where $p(g)$ is an estimate probability density function for continous gray levels $g$ and $h$ describes the density resolution.

The reasoning for the correctness of this metric is based on the assumption that the reconstructed image consists of homgenious regions, which is at least true for the outer regions that are zero.
In a reconstruction with wrong COR, additional structures are introduced which smear out the histogram.

\section{Conclusion}
In this paper about the determination of the COR in CT, a brief introduction was given about image acquisition in CT in the beginning~(\cref{ssec:imageacquisition}).
Several CT geometries were introduced; 
the parallel beam geometry~(\cref{sssec:parallelbeam}) which is mathematically described by the Radon transform and which most of the COR determination algorithms assume;
the fan beam geometry (\cref{sssec:fanbeam}) which is more common and can be transormed to a parallel beam geometry;
and the cone beam geometry (\cref{sssec:conebeam}) which, in contrast to the former, is a 3D geometry.
In the last part of the introduction to CT imaging, the reconstruction was discussed (\cref{ssec:reconstruction}) in which the Fourier slice theorem was explained as well as the filtered back projection (\cref{sssec:fbp}) as a representative reconstruction algorithm.

In the following section the key topic of this paper was adressed.
Firstly the problem of inaccurate or wrong COR assumption for reconstruction was explained (\cref{sec:cordetermination}) which is the image degradation due to artifacts.
Afterwards the different methdods for COR determination found in the literature were presented which can be classified into sinogram based methods and reconstruction based methods.

>>>>

% The paper headers
\markboth{Seminar on Image Processing - Algorithms and Architectures - University of Stuttgart - IPVS/PAS - SS17}%
{H{\"a}gele: Determination of Center of Rotation in Computed Tomography}
% The only time the second header will appear is for the odd numbered pages
% after the title page when using the twoside option.
% 
% *** Note that you probably will NOT want to include the author's ***
% *** name in the headers of peer review papers.                   ***
% You can use \ifCLASSOPTIONpeerreview for conditional compilation here if
% you desire.

@@

% for Computer Society papers, we must declare the abstract and index terms
% PRIOR to the title within the \IEEEtitleabstractindextext IEEEtran
% command as these need to go into the title area created by \maketitle.
% As a general rule, do not put math, special symbols or citations
% in the abstract or keywords.
\IEEEtitleabstractindextext{%
\begin{abstract}
This paper will discuss the aspect of the determination of center of rotation in computed tomography.
An introduction to computed tomography imaging will be given in the beginning as a prerequisite to the actual topic. 
This will cover different geometries (parallel beam, fan beam and cone beam), which are present in computed tomography and used to model the projection space of a CT-scan, as well as basic knowledge about tomographic reconstruction, including the Fourier slice theorem and filtered back projection.
The problem of wrong center of rotation assumption in tomographic reconstruction will be explained, which has the effect of image degradation.
Different solutions to the determination of the correct location of the center of rotation found in the literature will be presented.
It is found that the approaches can be split into sinogram and reconstruction based methods, where sinogram based methods analyze the sinogram that is the product of a CT-scan, while reconstruction based methods analyze reconstructed images of different centers of rotation to find the correct one.
Finally a conclusion summarizing the paper will be drawn.  
\end{abstract}

@@

% The very first letter is a 2 line initial drop letter followed
% by the rest of the first word in caps (small caps for compsoc).
% 
% form to use if the first word consists of a single letter:
% \IEEEPARstart{A}{demo} file is ....
% 
% form to use if you need the single drop letter followed by
% normal text (unknown if ever used by the IEEE):
% \IEEEPARstart{A}{}demo file is ....
% 
% Some journals put the first two words in caps:
% \IEEEPARstart{T}{his demo} file is ....
% 
% Here we have the typical use of a "T" for an initial drop letter
% and "HIS" in caps to complete the first word.
\IEEEPARstart{I}{n} computed tomography (CT), a set of set of projections is acquired by meassuring the attenuation of \xrays through the scanned object at different rotation angles.
The \xrays are emmited by a source on one side of the scanned object, intersect with the object, and hit a detector on the other side.
The resulting measurement of the detector is a projection of the object's internals.
Rotating the object and repeating the process results in a set of projections, which can be used to reconstruct an image of the internal structure of the scanned object. 
For this reconstruction it is mandatory to know the center of rotation (COR), or rather the location of its projection, as slight deviations from the actual COR produce severe artifacts in the reconstructed image.
In general, the COR is assumed to be projected onto the center of the detector, but this assumption can be violated on badly calibrated CT systems. 
A correct calibration can also become almost impossible when the \xray source is very close to the scanned object, as is the case for industrial scanners with resolutions on the micrometer scale \cite{azevedo90}.
In this paper, different techniques and algorithms for the determination of the center of rotation are explained, which can be found in the literature.
% You must have at least 2 lines in the paragraph with the drop letter
% (should never be an

@@

\section{Prerequisites}
In order to explain the problem of inaccurate COR assumption in tomographic reconstruction, some prerequisite knowledge about CT will be briefly explained in this section.

\subsection{Image Acquisition in CT}\label{ssec:imageacquisition}
This subsection will briefly explain how images are typically acquired in CT.

@@

\subsubsection{Parallel Beam}\label{sssec:parallelbeam}
The simplest geometry in CT (from a mathematical point of view) is the parallel beam geometry.
Here all rays are parallel that are emitted from the same side of the object.
\Cref{subfig:parallelbeam} shows a schematic of this geometry with the resulting signal.
%
\begin{figure}[!h]
\centering
\subfloat[Parallel Beam]{\includegraphics[width=0.58\linewidth]{img/parallel_beam_geometry}%
\label{subfig:parallelbeam}}
\hfil
\subfloat[Fan Beam]{\includegraphics[width=0.4\linewidth, trim=0 -6cm 0 0]{img/fanbeamgeometry}%
\label{subfig:fanbeam}}
\caption{Parallel beam and fan beam geometry. 
In parallel beam the source moving perpendicular to the ray direction for sampling a 1-dimensional image (graph). 
In fan beam, rays are not perpendicular but instead emitted in fan blade form. 
$\gamma$ is the rotation angle at which the signal is obtained. 
Figures from~\cite{Buzug2008_chap3,Buzug2008_chap4}}
\label{fig:parallalandfanbeam}
\end{figure}

@@

Unfortuantely, parallel beam geometry is impractical as a pencil beam \xray source (source emitting only one ray in a specific direction) would have to be moved parallel to the detector to collect all samples for a single rotation angle, which is very time consuming \cite{Buzug2008_chap7}.
However, parallel beam geometry occurs at synchrotron facilities \cite{something}.

\subsubsection{Fan Beam}\label{sssec:fanbeam}
A more practical geometry is the fan beam, in which a source emits a flat beam that extends to the sides (like a fan blade). 
\Cref{subfig:fanbeam} shows a schematic of this geometry.
Here the projection for a single rotation angle can be obtained at once. 
The geometry can be defined by the COR it rotates around, the offset of the source to the COR and its aperture angle.
In clinical practice curved detector rays are used, as shown in \cref{subfig:fanbeam}, to get constant aperture angle steps for each detector cell.
%
\begin{figure}[!h]
\centering
\subfloat[]{\includegraphics[width=0.43\linewidth]{eps/rebinning.eps}%
\label{subfig:rebinning1}}
\hfil
\subfloat[]{\includegraphics[width=0.51\linewidth]{eps/rebinning_geo.eps}%
\label{subfig:rebinning2}}
\caption{
On the left a schematic is shown with two fan beams with aperture angle $\varphi$ at different rotation angles.
It can be sen that the outer rays of the two are parallel.
The maximum distance of two parallel beams is limited by $\varphi$, and the maximum rotation angle $\theta$ at which two
parallel rays can be obtained is given by $\theta = \psi_{min}+\psi_{max}$ which are the angles from central to outer rays.
On the right a schematic is shown that points out the correspondence between fan and parallel beam geometry.
It can be seen that for the red ray which is the ray at rotation angle $\theta$ with deviation angle $\psi$ from the central ray, there exists an angle $\gamma$ in parallel beam geometry so that it is perpendicular to the $r$-axis (i.e. parallel beam detector).
Figures from~\cite{Buzug2008_chap7}
}
\label{fig:rebinning}
\end{figure}
%

@@

As th Radon transform is a set of projections the Fourier slice theorem can be used to build the Fourier tranform of the original signal and reconstruct it using the inverse Fourier transform.
\Cref{fig:inverseradon} illustrates this method.
%
\begin{figure}[!h]
\centering
\includegraphics[width=\linewidth]{img/inverseradon.png}
\caption{
The inverse Radon transform can be calculated using the Fourier slice theorem.
First all projections are transformed to the Fourier space along the $\xi$-axis ($=r$-axis).
Then the Fourier transform is built by placing all transformed projections according to their angle into a polar coordinate system.
The coordinate system thus consists of slices of the Fourier transform of the original signal which can then
be obtained using an inverse Fourier transform.
Figure from \cite{Buzug2008_chap5}
}
\label{fig:inverseradon}
\end{figure}
%
The problem of applying this algorithm in practice is, that due to the limited angular resolution (i.e. discretization) of the Radon transform the frequency information density in the Fourier space decreases for higher frequencies \cite{Buzug2008_chap5}.
In the process of cartesian regridding of the polar coordinates a lot of data for the high frequencies would have to be made up in order to perform an inverse fast Fourier transform.
Therefore a different algorithm, the filtered back projection, is used in practice.

@@

This can be imagined as smearing the projections over a canvas in their respective directions.
At first glance this may seem as the correct inversion of the projections but is not.
This becomes clear when thinking about the original image being a Dirac delta (image with all entries zero except for the central pixel).
The projections of this image will be 1D Dirac deltas, but using the back projection algorithm will not
restore the 2D Dirac but instead smear the 1D Diracs also into areas which are supposed to be zero.
In fact the resulting image looks quite blurry which is due to the back projection equaling to the original image convolved with the point spread function of the imaging system (i.e. the Radon transform) $\frac{1}{|(x,y)|}$ \cite{Buzug2008_chap5}.
To counter this, the projections first need to be filtered with a high pass filter.
This introduces negative values in the projections which in the back projection process cancel out positive contributions of other projections.
Figure \cref{fig:backprojection} shows the results of unfiltered and filtered back projection.
As amplifying high frequencies also amplifies noise, there are different filters that can be used in practice \cite{farquhar1998}.
%
\begin{figure}[!h]
\centering
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_02}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_04}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_08}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_16}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/bp_32}}
\\
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_02}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_04}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_08}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_16}}
\hfil
\subfloat{\includegraphics[width=0.18\linewidth]{img/fbp_32}}
\caption{
On the top row, the results of unfiltered back projection are shown. 
From left to right the number of projections used increases from 2 to 32.
The resulting image looks blurry.
On the bottom row the results of filtered back projection are shown.
The resulting image looks quite crips and edges are clearly visible.
The Shepp Logan phantom~\cite{shepplogan} was used as source image.  
}
\label{fig:backprojection}
\end{figure}
%

@@

\subsection{Sinogram Based Methods}
In this subsection the approaches that are based on sinogram analysis are introduced.
The approaches are presented in chronological order.

@@

This approach could also be expanded to use all projections in the sinogram.
For each opposing pair of projections the COR could be calculated and then the mean of all CORs will yield the final result.
This approach can be automated if there exists a distinct global maximum in the projections of the sinogram, otherwise a location of the correct feature is not guaranteed.

@@

Azevedo~et~al. showed that their algorithm is quite robust against Gaussian noise and isolated singularities~\cite{azevedo90}.
Unfortunately their algorithm is not applicable to fan beam geometry in general the objects center of mass is no longer projected onto the projections center of mass as was the case for parallel beam.
However for an object function $f(x,y) = \delta(x-\overline{x}, y-\overline{y})$, that is a Dirac delta function translated by $(\overline{x},\overline{y})$, the center of mass is projected onto the projections center of mass.
Therefore this method can be used for calibration (i.e. finding the COR beforehand) using an object that approximates a Dirac delta like a thin wire or pin phantom.

\subsubsection{Cross-Correlation}\label{sssec:crosscorrelation}
Yang et al.~\cite{Yang2011} proposed another sinogram based method for COR determination using image cross-correlation.
Given a sinogram in a parallel beam geometry over $360^\circ$ (which can also be obtained by rebinning a fan-beam sinogram), the sinogram is divided in half so that both halfs cover a $180^\circ$ range.
As the information in both sub sinograms is redundant in that way that they are mirror images of each other,
flipping one of them will result in two sinograms over $180^\circ$ with identical information.
%
\begin{equation}
R(r,\gamma)^{[0,\pi]} = R(-r,\gamma)^{[\pi,2\pi]}
\end{equation}
%
A shift of COR, as explained in the beginning of \cref{sec:cordetermination}, results in a shift of the sinogram along the $r$-axis (see~\cref{eq:radonoffset}).
Thus, the two sub sinograms are identical under this shift.
%
\begin{equation}
R_\tau(r+\tau,\gamma)^{[0,\pi]} = R_\tau(-(r+\tau),\gamma)^{[\pi,2\pi]}
\end{equation}
%
Using the 1D cross correlation of the two sinograms along the $r$-axis, the shift can be determined as follows.
%
\begin{equation}
2\tau = \underset{r}{arg\,max}\quad R_\tau(r,\gamma)^{[0,\pi]} \star R_\tau(-r,\gamma)^{[\pi,2\pi]}
\end{equation}
%
The authors~\cite{Yang2011} provided evidence that their algorithm is robust against Gaussian noise even for large variance values.
They also stated that their algorithm can be used for 3D CT with cone beam geometry when using the central slice (see \cref{sssec:conebeam}).
As locating the central slice is not trivial and depends on calibration as well, it was shown that the algorithm still performs well for sinograms that slightly deviate from that of the central slice.

@@

Recalling the filtering step in FBP where negative projection values are introduced to cancel out positive contribution (see~\cref{sssec:fbp}), it makes sense to analyze the negativity of the reconstructed image.
In a perfect reconstruction the negativity would be zero as the original image had no negative values, so minimizing this metric yields the correct COR.

@@

\subsubsection{Histogram Entropy}\label{sssec:qh}
As pointed out by Donath et~al.~\cite{donath2006}, the metric $Q_{IN}$ can only be used for non-negative object functions.
Therefore they provided another metric based on the histogram entropy that is also applicable when negative attenuation coefficients are present (which can occur when samples are recorded in a surrounding medium e.g. water \cite{donath2006}).
They defined a metric $Q_H$ as follows.
%
\begin{align}
Q_H(\tilde{f}) &= \frac{H}{H_{max}} 
\\
H &= -\int_{-\infty}^{\infty}p(g)\,log_2(h\,p(g))\;dg
\end{align}
%
Here the metric is build by normalizing the entropy $H$ by the maximum entropy $H_{max}$.
$H$ is a continuous version of the discrete entropy for gray-level images, where $p(g)$ is an estimate probability density function for continuous gray levels $g$ and $h$ describes the density resolution.

The reasoning for the correctness of this metric is based on the assumption that the reconstructed image consists of homogenious regions, which is at least true for the outer regions that are zero.
In a reconstruction with wrong COR, additional structures are introduced which smear out the histogram.

\section{Conclusion}
In this paper about the determination of the COR in CT, a brief introduction was given about image acquisition in CT in the beginning~(\cref{ssec:imageacquisition}).
Several CT geometries were introduced; 
the parallel beam geometry~(\cref{sssec:parallelbeam}) which is mathematically described by the Radon transform and which most of the COR determination algorithms assume;
the fan beam geometry (\cref{sssec:fanbeam}) which is more common and can be transformed to a parallel beam geometry;
and the cone beam geometry (\cref{sssec:conebeam}) which, in contrast to the former, is a 3D geometry.
In the last part of the introduction to CT imaging, the reconstruction was discussed (\cref{ssec:reconstruction}) in which the Fourier slice theorem was explained as well as the filtered back projection (\cref{sssec:fbp}) as a representative reconstruction algorithm.

In the following section the key topic of this paper was addressed.
Firstly the problem of inaccurate or wrong COR assumption for reconstruction was explained (\cref{sec:cordetermination}) which is the image degradation due to artifacts.
Afterwards the different methods for COR determination found in the literature were presented which can be classified into sinogram based methods and reconstruction based methods.
